---
title: "Assignment 5.2"
author: "Mario"
date: "11 Dezember 2016"
output: md_document # html_document alternatively
---

```{r setup, include=TRUE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(lubridate)
library(dplyr)
library(zoo)
library(ggplot2)
options(scipen=999) # no scientific notation
```

## Loading and preprocessing the data

```{r getdata}
# 1. Load the data
mydata <- read.csv(file='activity.csv')
head(mydata)
str(mydata)

# 2. Process / transform the data
mydata$date <- ymd(mydata$date) # ymd from package lubridate
str(mydata)
```

## What is the mean total number of steps taken per day?

```{r figure_1_steps}
# 1. Calculate the total number of steps taken per day
mysteps <- mydata %>% 
    group_by(date) %>% 
    summarise(sumSteps = sum(steps))
mysteps 

# 2. Make a histogram
hist(x = mysteps$sumSteps, main = 'Total number of daily steps for 61 days', 
     xlab = 'Number of steps per day')
 
# 3. Calculate and report the mean and median
stepsMean <- round(mean(mysteps$sumSteps, na.rm = TRUE), digits = 1)
stepsMedian <- round(median(mysteps$sumSteps, na.rm = TRUE), digits = 1)
```
The mean of the total number of steps taken per day is **`r stepsMean`**, 
the median is **`r stepsMedian`**.

## What is the average daily activity pattern?

```{r figure_2_steps}
# 1. Make a time series plot of the 5-minute interval and the average number of steps taken
mysteps2 <- mydata %>% 
    group_by(interval) %>% 
    summarise(meanSteps = mean(steps, na.rm = TRUE))
plot(x = mysteps2$interval, y = mysteps2$meanSteps, type = 'l',
     ylab = 'Average number of steps', xlab = 'Interval',
     main = 'Average number of interval steps for 61 days')

# 2. Which interval contains the maximum number of steps?
maxInterval <- mysteps2[mysteps2$meanSteps == max(mysteps2$meanSteps),][1]
```

The interval with the maximum number of steps is **`r maxInterval`**.

## Imputing missing values

```{r complete}
# 1. Calculate & report the number of rows with NAs
mydata <- mutate(mydata, complete = complete.cases(mydata)) # add variable that indicates completeness
complete <- sum(mydata$complete == TRUE)
incomplete <- sum(mydata$complete == FALSE)
pincomplete <- round(incomplete/(complete + incomplete), digits = 3)
# 2. Devise a strategy for filling in all of the missing values in the dataset
```
In `r complete + incomplete` cases there are **`r incomplete`** cases with NAs (p = `r pincomplete`).

Missing values will be replaced by average steps of the respective interval.

```{r figure_3_missingsFilled}
# 3. Create a new dataset with missing data filled in
mydata2 <- mydata
head(mydata2) 
# library zoo => na.aggregate()
mydata2$steps <- na.aggregate(mydata2$steps, by = mydata2$interval, na.rm = TRUE)
head(mydata2) 

# 4. Repeat histogram, mean, and median total number of steps taken per day
mysteps2 <- mydata2 %>% 
    group_by(date) %>% 
    summarise(sumSteps = sum(steps))
hist(x = mysteps2$sumSteps, main = 'Total number of daily steps for 61 days\n(missing values replaced by interval mean)', 
     xlab = 'Number of steps per day')
stepsMean2 <- round(mean(mysteps2$sumSteps, na.rm = TRUE), digits = 1)
stepsMedian2 <- round(median(mysteps2$sumSteps, na.rm = TRUE), digits = 1)
```

The mean of the total number of steps taken per day is now **`r stepsMean2`**, 
the median is **`r stepsMedian2`**. This means that the impact of filling missing values 
can be seen with regard to absolute frequencies in the histogram 
but not that much with respect to the mean or median of the number of steps taken.
That being said, the median approaches the mean a bit more.


## Are there differences in activity patterns between weekdays and weekends?

```{r figure_4_weekend}
# 1. Create factor variable with levels 'weekday' and 'weekend'
# lubridate::wday being used
mydata2 <- mutate(mydata2, dayposition = ifelse(wday(date) %in% c(1,7),'weekend', 'weekday'))
with(mydata2, table(dayposition))

# 2. Make a panel plot containing a time series of interval on x and the average number of steps taken on y
# => 1 plot for weekdays and 1 plot for weekdays

paneltable <- mydata2 %>% 
    group_by(dayposition, interval) %>% 
    summarise(averageSteps = mean(steps))
paneltable$interval <- as.numeric(paneltable$interval)

# ggplot2 package being used
ggplot(paneltable, aes(interval, averageSteps)) + geom_line(colour = 'blue') +
 facet_wrap(~dayposition, nrow=2) +  
 ggtitle('Average Steps taken in consecutive time intervals of a day\n- Comparison of Weekdays and weekends')
```
